Lossy compression of DNA sequencing quality data

### 2015-02-04 manuscript/q-compression.tex
More revisions to remove from the results that material that more appropriate belongs in the intro and methods sections. Added bonus: methods section is typeset in a smaller font and thus leaves more room for other sections.

\copyrightyear{2014}
\pubyear{2014}

@@

Quality values are typically used throughout bioinformatics pipelines.
Software such as \textsc{fastqc}~\cite{citeulike:11583827} and
Prinseq~\cite{Schmieder:2011gd} provide the user with detailed
descriptions of the distribution of sequence quality scores, and these
scores can be used to guide filtering and trimming of sequence read
data as processing steps before subsequent analyses. Quality values
can be used by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

@@

\subsection{Binning}

Quality values can be binned into a set of bins, and any number of
bins less than the the number of distinct quality values is
possible. The minimum number of bins that allows for a any distinction
among quality values is two, i.e., two categories ``good'' and ``bad''
quality. We implement 2-bin encoding by setting a quality value
threshold empirically determined by the distribution of quality values
across reads. Base-calls are marked ``bad'' if their quality value
falls below the first quartile minus 1.5 $\times$ the interquartile
range (IQR), which is the difference between the first and third
quartile. 1.5 $\times$ IQR is the value used by Tukey's box
plot~\cite{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent. No prior assumptions of the
distribution of the quality values need to be made.

@@

\subsection{Modeling}

@@

\subsection{Profiling}

@@

As a measure of compression effectiveness we use bits/base-call, and
define it as the size of the compressed representation of quality
values (in bits) divided by the number of quality values represented.
As a measure of information loss we use mean squared error
(\textsc{mse}) as aloss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
sequences, $Q_i$ is the original quality value associated with
sequence position $i$, and $Q_i'$ is the corresponding
compressed/decompressed quality value.

For quality control, we use
Sickle~\cite{sickle}. PrinSeq~\cite{Schmieder:2011gd} was used to
count the number of bases and reads in each \textsc{fastq} file.

For \emph{de novo} genome assembly, we use
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

For aligning the sequences, we use Bowtie2~\cite{Langmead:2012rw} with
default settings and modified max- and min-mismatch parameters
detailed below.

@@

\subsection{Compressibility versus error}

@@

One way to evaluate different compression approaches is to perform
quality preprocessing on the generated \textsc{fastq} files and
compare the number of reads and bases with those of the uncompressed
\textsc{fastq} files. Sickle is a tool to filter and trim reads. We
used PrinSeq to calculate the number of reads and bases in each
sample. Table \ref{tab:quality_control} shows the proportional numbers
with respect to those of the uncompressed \textsc{fastq} files. In the
fragment sample, we retain on average 95.4\% and 94.24\% of bases and
reads, respectively. In the short-jump sample, we retain 90.4\% and
90.98\% of bases and reads, respectively. In some cases, the
proportion of reads and/or bases compared to the original is more than
100\%; this means that the number of the reads and bases in the
compressed samples are more because altering quality values may
prevent the read from being filtered and/or trimmed. This is clear in
the good/ bad compression approach of the short-jump sample, where the
proportion of bases and reads are 113.09\% and 103.70\% respectively.

@@

Genome assemblers, such as \textsc{allpaths-lg}~\cite{Gnerre:2011kx},
often utilize quality values to aid in the assembly process. By
reducing the information contained within the quality values, we
introduce the potential for the resulting assembly to be affected.

\textsc{allpaths-lg} is one of the leading bacterial genome
assemblers~\cite{Salzberg:2012rc}. We examine the effect of our lossy
compression methods on the \textit{Rhodobacter sphaeroides} 2.4.1 data
set provided by \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}. \textsc{allpaths-lg} is used to
assemble the original data along with the quality compressed versions
using 2-bin encoding, 0--5th-order polynomial regression encodings,
and 32- and 256-profile encodings. We evaluate the assembly based on
contiguity statistics, log average read probability
(\textsc{lap}~\cite{Ghodsi:2013hb}), and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} which counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. Alongside the raw metrics, we
also provide a cummulative sum of the individual ranks of each method
for each metric. A similar approach was utilized by the
\textsc{Assemblathon1} competition~\cite{Earl:2011fv}.

The assembly using the uncompressed reads outperforms all compression
methods in terms of \textsc{lap}, NG50, relocations, translations, and
inversions (Table \ref{tab:assembly_ranks}). Among the compression
methods, the 256- and 32-profile encoding had the highest rank, then
the 4th-order polynomial regression, the 5th-order polynomial
regression, the 2-bin encoding, and lastly, the 3rd through 0-order
polynomials.

@@

As described previously, quality values in Illumina reads are
non-uniformly distributed over the read length and show a general
pattern of decreasing toward the 3$^{\prime}$ end. As large sets of
quality strings show similar trends of quality over their sequence,
one approach would be to identify common patterns in the data,
generate profiles representing these patterns, and use the profiles as
proxies for read quality value sequences.

@@

\paragraph{Read quality processing}
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases.

To evaluate the effect of compression on these quality control
operations, we used Sickle~\cite{sickle} with the default
settings. \textsc{PrinSeq}~\cite{Schmieder:2011gd} was then used to
count the number of bases and reads in each quality control-processed
\textsc{fastq} file. The proportion of reads retained using
compressed/decompressed data compared to that of the uncompressed
data: 0.998 for the fragment library reads, and 1.013 for the
short-jump library reads; and proportion of bases retained 1.0 for the
fragment library reads, and 1.027 for the short-jump library
reads. Thus there is very little difference between profile-based
compressed data compared to that of the original uncompressed data in
read filtering and trimming.

\paragraph{\textit{De novo} genome assembly}
For de novo genome assembly, we use
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings. We evaluate the assembly based on contiguity statistics, log
average read probability (\textsc{lap})~\cite{Ghodsi:2013hb}, and a
collection of reference-based metrics, of which the contiguity
statistics include number of scaffolds and NG50; the latter statistic
is defined as the weighted median contig size (the length of largest
contig $c$ such that the total size of the contigs larger than $c$
exceeds half of the known genome size). The \textsc{lap} score can be
viewed as a log likelihood score, where a larger value is better. We
use a reference-based evaluation script provided by \textsc{gage},
which counts single nucleotide polymorphisms (\textsc{snp}s),
relocations, translations, and inversions. The reference- based
metrics are normalized by the length of the assembly to aid in
comparison. The differences in assembly statistics between using the
profile-based compressed/decompressed data compared to using
uncompressed data are extremely small: total assembly size difference
of 6.745 kbp ($\sim$0.00015); NG50 size difference of 2.259 kbp
(0.0007); an \textsc{lap} score difference of 0.002 for the fragment
library and 0.04 for the short-jump library.

>>>>

\copyrightyear{2015}
\pubyear{2015}

@@

Quality values are typically used throughout bioinformatics pipelines.
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

@@

\subsection{Compression strategies}

\subsubsection{Binning}

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two, i.e., two
categories ``good'' and ``bad'' quality. We implement 2-bin encoding
by setting a quality value threshold empirically determined by the
distribution of quality values across reads. Base-calls are marked
``bad'' if their quality value falls below the first quartile minus
1.5 $\times$ the interquartile range (IQR), which is the difference
between the first and third quartile. 1.5 $\times$ IQR is the value
used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
of this approach is that it is completely data-dependent. No prior
assumptions of the distribution of the quality values need to be made.

@@

\subsubsection{Modeling}

@@

\subsubsection{Profiling}

@@

As a measure of compression effectiveness we use bits/base-call, and
define it as the size of the compressed representation of quality
values (in bits) divided by the number of quality values represented.
As a measure of information loss we use mean squared error
(\textsc{mse}) as aloss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
sequences, $Q_i'$ is the compressed/decompressed quality value, and
$Q_i$ is the original quality value associated with sequence position
$i$.

We evaluate effects of information loss from quality value compression
on quality control steps of read filtering and trimming, which were
performed using Sickle~\cite{sickle}, quantify read characteristics
with PrinSeq~\cite{Schmieder:2011gd}, and make comparison to
uncompressed data.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include number of
scaffolds and NG50, which is defined as the weighted median contig
size (the length of largest contig $c$ such that the total size of the
contigs larger than $c$â exceeds half of the known genome size). The
\textsc{lap} score can be viewed as a log likelihood score, where a
value closer to 0 is better. We use a reference-based evaluation
script provided by \textsc{gage} that counts single nucleotide
polymorphisms (\textsc{snp}s), relocations, translations, and
inversions. The reference-based metrics are normalized by the length
of the assembly to aid in comparison. For the genome assembly we used
software that makes use quality values in the assembly process:
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

\subsection{Compression effectiveness versus information loss}

@@

Table \ref{tab:quality_control} shows the proportional numbers with
respect to those of the uncompressed \textsc{fastq} files. In the
fragment sample, we retain on average 95.4\% and 94.24\% of bases and
reads, respectively. In the short-jump sample, we retain 90.4\% and
90.98\% of bases and reads, respectively. In some cases, the
proportion of reads and/or bases compared to the original is more than
100\%; this means that the number of the reads and bases in the
compressed samples are more because altering quality values may
prevent the read from being filtered and/or trimmed. This is clear in
the good/ bad compression approach of the short-jump sample, where the
proportion of bases and reads are 113.09\% and 103.70\% respectively.

@@

The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
using the uncompressed reads outperforms all compression methods in
terms of \textsc{lap}, NG50, relocations, translations, and inversions
(Table \ref{tab:assembly_ranks}). Among the compression methods, the
256- and 32-profile encoding had the highest rank, then the 4th-order
polynomial regression, the 5th-order polynomial regression, the 2-bin
encoding, and lastly, the 3rd through 0-order polynomials.

@@

@@
